---
title: "p8105_hw2_fp2513"
output: github_document
---

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
library(haven)
```

##Problem 1

**Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or case_match function may be useful).**

```{r}
NYC_Transit_df = 
  read_csv(file = "NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
           na = c("NA", ",", ".")) %>% 
  janitor::clean_names() %>% 
  select(line:entry, vending, ada) %>% 
  mutate(
    entry = 
      case_match(
        entry, 
        "NO" ~ FALSE, 
        "YES" ~ TRUE),
    entry = as.logical(entry))

```

**Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?**

After importing and intial cleaning of the data, the NYC_Transit_df contains 1868 observations (each having its own row; 1868 rows) and 19 variables (19 columns). The data cleaning steps so far include:
1.    First defining the cells with NA, . and empty cells are used to indicate a missing value. 
2.    Then using `janitor::clean_names()` to clean up the variable column names to a consistent lower snake case. 
3.    Next, `select()` specified the columns that I wanted to keep, including the range from line to entry, and additional vending and ada columns
4.    Lastly, `mutate()` converted the inputs in the entry column from characters to logical variables. Specifically, using `case_match` to chnage every "YES" and "NO" input respectively became "TRUE" and "FALSE" instead.  

The data is clean in respects to ensuring that every variable have their own column, every observation has its own row and every box / input is just some combination of the variable and observation. The data is however not clean in respects to redundant columns. There are 11 columns used for route 1 to 11 where as 'Route number' should be a distinct column for each observation. There should also be an additional column called 'Route name'. (Specify the route number then which the name of that route is in two columns). 

After cleaning the data according to the instructions, there are also rows of identical data inputs. We could consider evaluating the duplicates in a new column to indicate how many times it was repeated. 
